---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
options(width = 80)
library(Rducks)
```

# Rducks

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rducks/actions/workflows/R-CMD-check.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rducks)](https://sounkou-bioinfo.r-universe.dev/Rducks)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

Rducks registers R functions as DuckDB SQL functions using a package-managed
[DuckDB C extension](https://duckdb.org/docs/stable/extensions/overview). The
extension records the DuckDB database instance at load time and keeps
extension-owned connections for native registration callbacks. It is built
around explicit type descriptors, direct DuckDB-vector marshalling, and a strict
rule that R object work runs on the recorded R thread.

Rducks is organized around DuckDB function kind, scalar-UDF evaluation mode, and
execution plan. Scalar UDFs are registered with `rducks_register_scalar_udf()`
and choose `mode = "scalar"` for one R call per row or `mode = "vectorized"` for
one R call per DuckDB chunk. Two execution transports are available:
in-process (`transport = "inproc"`) and worker-process (`transport = "ipc"`,
native [NNG](https://nng.nanomsg.org/) request/reply, default local workers
launched by [mirai](https://mirai.r-lib.org/), optional
[mori](https://shikokuchuo.net/mori/) sharing for selected globals). Aggregates
use `rducks_register_aggregate()`.

## How it works

Rducks loads a small DuckDB extension that records the database instance and
keeps extension-owned connections for registration callbacks. R closures are
preserved while DuckDB catalog metadata can call them, and scalar calls run on
the recorded R thread or, under `transport = "ipc"`, in worker R processes.

In-process marshalling is direct: DuckDB produces vectors in standard chunks;
Rducks materializes typed R inputs directly from those vectors in extension C,
calls the R function, and writes typed results back into DuckDB output vectors.
When a callback arrives off the recorded R thread under the in-process concurrent
backend, the input chunk and the result are carried as owned DuckDB data chunks
across the queue boundary. Dynamic omitted-`args` UDFs still bind concrete DuckDB
types at the SQL call site before this marshalling begins.

The worker-process transport uses a
[Quack-style format](https://github.com/duckdb/duckdb-quack): DuckDB
`BinarySerializer` messages carrying logical types and `DataChunk` payloads,
aligned with DuckDB's native chunk model. The codec
lives in `src/quack_core.c` with R glue in `src/quack_codec.c`. The worker path
currently marshals fixed-width scalars, `VARCHAR`/`BLOB`, `DECIMAL`, `INTERVAL`,
`ENUM`, `BIT`, `GEOMETRY`, `MAP`, `UNION`, and `LIST`/`ARRAY`/`STRUCT` of
supported types; `VARIANT` is rejected at registration on the `ipc` plan until
the native bridge covers it.

## Quick start

```{r quick-start}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")

score_udf <- rducks_register_scalar_udf(
  con,
  name = "r_score",
  fun = function(row) {
    bonus <- if (identical(row$label, "high")) 100 else 0
    list(
      score = as.double(row$x + bonus),
      parts = as.double(c(row$x, bonus))
    )
  },
  returns = STRUCT(score = DOUBLE, parts = DOUBLE[]),
  side_effects = TRUE
)

dbGetQuery(con, "
  WITH input AS (
    SELECT struct_pack(x := x::DOUBLE, label := label) AS payload
    FROM (VALUES (2, 'low'), (21, 'high')) AS t(x, label)
  ), scored AS (
    SELECT r_score(payload) AS result FROM input
  )
  SELECT result.score AS score, result.parts AS parts
  FROM scored
")
```

`r_score()` omits `args`, so DuckDB registers it as a dynamic varargs scalar
UDF. At this SQL call site DuckDB binds a concrete `STRUCT(x DOUBLE, label
VARCHAR)` input, and the return type is explicit: a struct containing a `DOUBLE`
and a `DOUBLE[]`. Rducks materializes dynamic inputs as if the signature had been
declared with `args = ...`. Use `args = NULL` only for a true zero-argument UDF.

The returned registration object records the normalized signature and options;
DuckDB owns the catalog function after registration. Dropping the R object does
not unregister the SQL function. Registering the same SQL name/signature again
replaces the callable implementation. Use `side_effects = TRUE` for functions
with counters, randomness, I/O, mutation, sleeps, or other effects so DuckDB does
not optimize them as pure expressions.

## Lifecycle

`rducks_release(con)` detaches connection-local Rducks state. For file-backed
databases, it also closes Rducks' extension-owned DuckDB connections so the file
can be fully closed on strict file-locking platforms. It does not drop DuckDB
catalog functions or release closures still owned by native catalog metadata. For
deterministic cleanup, call it before `DBI::dbDisconnect(con)`; to replace a
scalar UDF, register the same SQL name/signature again.

## Type descriptors

Rducks descriptors are used for scalar-UDF returns, declared scalar-UDF inputs,
and aggregate inputs/returns. They include DuckDB scalar types, exact value
classes such as `UUID`, `HUGEINT`, `DECIMAL(width, scale)`, `INTERVAL`, `BIT`,
`GEOMETRY`, `VARIANT`, `ENUM(levels)`, and composite descriptors such as
`LIST(TYPE)`, `ARRAY(TYPE, n)`, `STRUCT(...)`, `MAP(key, value)`, and
`UNION(...)`. `GEOMETRY` values cross as their opaque physical `raw` bytes (WKB
for a real geometry); `VARIANT` values cross
as DuckDB's typed storage struct wrapped by `rducks_variant`. VARIANT scalar-UDF
signatures require a DuckDB runtime whose C API exposes VARIANT logical types and
are not supported by direct marshalling yet. Direct UNION support follows
DuckDB's native UNION vector tag/child layout; it is tested for supported DuckDB
builds but is intentionally treated as a version-coupled native adapter rather
than a stable interchange format.

```{r type-descriptors}
nested_type <- STRUCT(
  id = INTEGER,
  label = ENUM(c("low", "high")),
  payload = UNION(code = INTEGER, note = VARCHAR),
  values = LIST(DECIMAL(10, 2))
)

rducks_is_type(nested_type)
cat(strwrap(rducks_type_sql(nested_type), width = 70), sep = "\n")
rducks_type_child_names(nested_type)
```

## Scalar UDFs

Scalar mode calls the R function once per logical row. Vectorized mode calls the
R function once per DuckDB chunk with one R vector or list-column per declared or
dynamically bound argument.

```{r vectorized-scalar-udf}
scalar_plus_one_udf <- rducks_register_scalar_udf(
  con,
  name = "r_scalar_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "scalar",
  side_effects = TRUE
)

vec_plus_one_udf <- rducks_register_scalar_udf(
  con,
  name = "r_vec_plus_one",
  fun = function(x) x + 1,
  args = DOUBLE,
  returns = DOUBLE,
  mode = "vectorized",
  side_effects = TRUE
)

dbGetQuery(con, "SELECT sum(r_vec_plus_one(i::DOUBLE)) AS total FROM range(5) AS t(i)")
```

Dynamic omitted arguments are not a guessing path. They are bind-time descriptors.
The same R function below is registered once with an explicit nested signature
and once as dynamic varargs; both calls see the same typed R value.

```{r declared-vs-dynamic}
nested_summary <- function(x) {
  paste0(x$id, ":", x$label, ":", x$payload$tag, "=", x$payload$value)
}

nested_declared_udf <- rducks_register_scalar_udf(
  con,
  name = "r_nested_declared",
  fun = nested_summary,
  args = STRUCT(
    id = INTEGER,
    label = ENUM(c("low", "high")),
    payload = UNION(code = INTEGER, note = VARCHAR)
  ),
  returns = VARCHAR,
  null_handling = "special"
)

nested_dynamic_udf <- rducks_register_scalar_udf(
  con,
  name = "r_nested_dynamic",
  fun = nested_summary,
  returns = VARCHAR,
  null_handling = "special"
)

nested_sql <- "
  struct_pack(
    id := 7::INTEGER,
    label := 'high'::ENUM('low', 'high'),
    payload := union_value(note := 'ok')::UNION(code INTEGER, note VARCHAR)
  )
"

nested_query <- sprintf(
  paste(
    "SELECT",
    "  r_nested_declared(%1$s) AS declared,",
    "  r_nested_dynamic(%1$s) AS dynamic",
    sep = "\n"
  ),
  nested_sql
)
dbGetQuery(con, nested_query)
```

With `null_handling = "default"`, a top-level SQL `NULL` input produces a SQL
`NULL` output without calling R. With `null_handling = "special"`, top-level
SQL `NULL` values are passed as type-specific R missing values. Nested NULLs are
part of the value: scalar children usually become typed `NA`, while nested
composite NULLs become `NULL`.

```{r null-handling}
null_special_udf <- rducks_register_scalar_udf(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

For type-by-type details, use the exported reference tables:

```{r semantic-tables}
rducks_mode_semantics()[, c("mode", "call_granularity", "input_shape")]
mapping <- rducks_argument_type_mapping(list(
  INTEGER,
  UUID,
  DECIMAL(10, 2),
  STRUCT(a = INTEGER[])
))
mapping[, c("duckdb_type", "r_value_class", "special_null_argument")]
```

## Aggregates

`rducks_register_aggregate()` registers R-backed DuckDB aggregates. Aggregate
state is an R object preserved by the extension, not a serialized blob. The
callbacks run on the recorded R thread and are not controlled by scalar-UDF
execution plans.

```{r aggregate}
sum_i32_aggregate <- rducks_register_aggregate(
  con,
  name = "r_sum_i32",
  update = function(state, x) {
    if (is.null(state)) state <- 0L
    as.integer(state + x)
  },
  finalize = function(state) if (is.null(state)) NA_integer_ else state,
  args = INTEGER,
  returns = INTEGER
)

dbGetQuery(
  con,
  paste(
    "SELECT r_sum_i32(i) AS total",
    "FROM (VALUES (1::INTEGER), (2::INTEGER), (NULL::INTEGER)) t(i)"
  )
)
```

## Table functions

`rducks_register_table()` infers the number of SQL arguments from the R function
formals and registers those inputs as DuckDB `ANY`. The result can be a data
frame, a named list of columns, or `rducks_table_stream()` for scan-time batches.
Column types are inferred from the returned columns, and the extension fills
DuckDB output vectors directly from the R columns, with no wire serialization
for the in-process scan.

```{r table-function}
rows_table <- rducks_register_table(
  con,
  name = "r_rows",
  fun = function(n) data.frame(i = seq_len(as.integer(n))),
  chunk_size = 2L
)

dbGetQuery(con, "SELECT * FROM r_rows(3) ORDER BY i")

stream_rows_table <- rducks_register_table(
  con,
  name = "r_stream_rows",
  fun = function(n) {
    next_i <- 1L
    limit <- as.integer(n)
    rducks_table_stream(
      prototype = data.frame(i = integer()),
      next_batch = function(batch_size) {
        if (next_i > limit) return(NULL)
        hi <- min(limit, next_i + as.integer(batch_size) - 1L)
        out <- data.frame(i = seq.int(next_i, hi))
        next_i <<- hi + 1L
        out
      }
    )
  },
  chunk_size = 2L
)

dbGetQuery(con, "SELECT sum(i) AS total FROM r_stream_rows(5)")
```

## Query streams

`rducks_query_stream()` returns a query's rows in DuckDB-sized batches as data
frames, instead of an eager `DBI::dbGetQuery()` result. Each batch is
materialized directly from DuckDB vectors to R values on the recorded R thread.
`next_batch()` returns the next data frame or `NULL` at end of stream.

```{r query-stream}
stream <- rducks_query_stream(con, "SELECT i::INTEGER AS i FROM range(1, 6) t(i)")
stream$next_batch()
stream$next_batch()
stream$close()
```

Because each batch is one DuckDB chunk that R materializes and then releases,
memory stays bounded regardless of result size: the stream never holds the whole
result. The query below produces ~8 million rows with a padded column (well over
half a gigabyte if collected eagerly with `dbGetQuery()`), yet the resident set
stays a small, roughly constant fraction of that.

```{r query-stream-memory}
rss_mb <- function() {
  vm <- grep("VmRSS", tryCatch(readLines("/proc/self/status"), error = function(e) ""), value = TRUE)
  if (length(vm)) round(as.numeric(gsub("[^0-9]", "", vm)) / 1024) else NA_real_
}

stream <- rducks_query_stream(
  con,
  "SELECT i, i * i AS sq, repeat('x', 64) AS pad FROM range(8000000) t(i)"
)
rss_at_open <- rss_mb()
rows <- 0
peak <- rss_at_open
repeat {
  batch <- stream$next_batch()
  if (is.null(batch)) break
  rows <- rows + nrow(batch)          # consume one chunk, then let it be GC'd
  peak <- max(peak, rss_mb())
}
stream$close()
data.frame(rows = rows, rss_at_open_mb = rss_at_open, peak_rss_mb = peak)
```

The same streaming holds over arbitrary scans, including external table-function
extensions. The chunk below streams a 15 GB BAM through a `read_bam()` scanner
(the [duckhts](https://github.com/RGenomicsETL/duckhts) extension), capped at ~30
seconds of pulling; resident memory stays a few hundred MB no matter how many
reads are pulled. It runs only when the BAM is present locally.

```{r bam-stream, eval = file.exists("NA12878.low_coverage.bam")}
DBI::dbExecute(con, "LOAD './duckhts.duckdb_extension'")   # locally built duckhts

stream <- rducks_query_stream(
  con,
  "SELECT QNAME, FLAG, RNAME, POS, MAPQ, CIGAR, SEQ, QUAL
     FROM read_bam('NA12878.low_coverage.bam')"   # 15 GB on disk
)
bam_open <- rss_mb()
reads <- 0
peak <- bam_open
deadline <- Sys.time() + 30
repeat {
  batch <- stream$next_batch()
  if (is.null(batch) || Sys.time() > deadline) break
  reads <- reads + nrow(batch)          # one chunk at a time, then GC'd
  peak <- max(peak, rss_mb())
}
stream$close()
data.frame(
  bam_gb = round(file.size("NA12878.low_coverage.bam") / 1024^3, 1),
  reads_streamed = reads,
  rss_at_open_mb = bam_open,
  peak_rss_mb = peak
)
```

## Execution plans

Execution plans are fixed at scalar-UDF registration time and select the
placement/concurrency backend for the recorded R thread.

| Plan | Engine | Scalar mode | Vectorized mode | Notes |
| --- | --- | --- | --- | --- |
| `inproc` (`direct + inproc_concurrent`) | `direct_main_queue` | yes | yes | DuckDB workers enqueue callbacks and R work drains on the recorded R thread |
| `ipc` (`wire + multiprocess_parallel`) | `ipc_nng_pool` | yes | yes | the extension marshals each chunk to a worker R process over NNG using the Quack wire codec; fixed-width scalars, varchar/blob, decimal, interval, enum, bit, and list/array/struct of supported types |

`rducks_enable(con)` selects the in-process backend; `rducks_enable_inproc(con,
threads = 4, external_threads = 1)` keeps R work on the recorded R thread while
DuckDB workers run concurrently and queue their callbacks to it.

For `transport = "ipc"`, register single-threaded under the plan (this starts the
worker processes and broadcasts the UDF), then raise DuckDB threads so off-main
callbacks fan out to the workers:

``` r
plan <- rducks_execution_plan("ipc", ipc_workers = 2L)
rducks_set_execution_plan(con, plan, threads = 1L, external_threads = 1L)
rducks_register_scalar_udf(con, "r_par", function(x) x + 1L, args = INTEGER, returns = INTEGER)
rducks_set_execution_plan(con, plan, threads = 3L, external_threads = 2L)
dbGetQuery(con, "SELECT sum(r_par((i % 1000)::INTEGER)) AS total FROM range(20000) t(i)")
```

```{r execution-plan}
plan <- rducks_execution_plan("inproc")
plan

# Register single-threaded, then raise DuckDB threads for concurrent execution
# of the already-registered UDF; off-main worker callbacks queue to the
# recorded R thread.
rducks_register_scalar_udf(
  con,
  name = "r_inc",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER,
  mode = "vectorized"
)
rducks_set_execution_plan(con, plan, threads = 4L, external_threads = 1L)
dbGetQuery(con, "SELECT sum(r_inc((i % 1000)::INTEGER)) AS total FROM range(20000) t(i)")
rducks_set_execution_plan(con, plan, threads = 1L, external_threads = 1L)
```

### When worker-process execution wins

`inproc` runs every R callback on the single recorded R thread, so an expensive
per-chunk UDF serializes even while DuckDB scans in parallel. `transport = "ipc"`
fans those chunks out to worker R processes, so the per-chunk cost overlaps once
it is worth the marshalling. The benchmark registers the same sleeping
vectorized UDF on three plans and runs each against one large parallel CSV scan.
The UDF closes over a random R lookup vector; the third plan ships that global to
the workers through [mori](https://shikokuchuo.net/mori/) shared memory
(`ipc_globals_share = "mori"`). Timings are illustrative and machine-dependent.

```{r execution-plan-benchmark}
set.seed(1)
lookup <- sample.int(20L, 1000L, replace = TRUE)
slow_lookup <- function(x) {
  Sys.sleep(0.05)
  x + lookup[[1L]]
}

duckdb_vector_size <- 2048L
csv_rows <- duckdb_vector_size * 64L
csv_pad <- strrep("x", 128L)
csv_path <- tempfile("rducks-readme-csv-", fileext = ".csv")
writeLines(
  c("i,pad", paste0(seq.int(0L, csv_rows - 1L), ",", csv_pad)),
  csv_path
)

ipc_workers <- 2L
plans <- list(
  inproc = rducks_execution_plan("inproc"),
  ipc = rducks_execution_plan(
    "ipc",
    ipc_workers = ipc_workers,
    ipc_timeout = 60,
    ipc_globals = "lookup"
  ),
  ipc_mori = rducks_execution_plan(
    "ipc",
    ipc_workers = ipc_workers,
    ipc_timeout = 60,
    ipc_globals = "lookup",
    ipc_globals_share = "mori"
  )
)
udfs <- paste0("r_bench_", names(plans))

# Register each UDF single-threaded under its plan; for the ipc plans this
# starts the worker processes and broadcasts the UDF (and its globals).
for (i in seq_along(plans)) {
  rducks_set_execution_plan(con, plans[[i]], threads = 1L, external_threads = 1L)
  rducks_register_scalar_udf(
    con,
    name = udfs[[i]],
    fun = slow_lookup,
    args = INTEGER,
    returns = INTEGER,
    mode = "vectorized",
    side_effects = TRUE
  )
}

run_plan <- function(label, udf, plan, threads, external_threads) {
  rducks_set_execution_plan(con, plan, threads = threads, external_threads = external_threads)
  elapsed <- system.time({
    result <- DBI::dbGetQuery(con, sprintf(
      paste(
        "SELECT sum(%s((i %% 1000)::INTEGER)) AS total",
        "FROM read_csv(%s, header = true,",
        "columns = {'i': 'INTEGER', 'pad': 'VARCHAR'}, parallel = true)"
      ),
      DBI::dbQuoteIdentifier(con, udf),
      DBI::dbQuoteString(con, csv_path)
    ))
  })[["elapsed"]]
  data.frame(label = label, total = result$total[[1]], elapsed_sec = round(elapsed, 3))
}

benchmark <- rbind(
  run_plan("inproc (single R thread)", udfs[[1]], plans[[1]],
           threads = 1L, external_threads = 1L),
  run_plan("ipc (2 workers)", udfs[[2]], plans[[2]],
           threads = ipc_workers + 1L, external_threads = ipc_workers),
  run_plan("ipc + mori (2 workers)", udfs[[3]], plans[[3]],
           threads = ipc_workers + 1L, external_threads = ipc_workers)
)
unlink(csv_path, force = TRUE)
rducks_set_execution_plan(con, rducks_execution_plan("inproc"),
                          threads = 1L, external_threads = 1L)
benchmark
```

## duckplyr integration

`rducks_with_duckplyr()` and the `with.duckdb_connection()` method let ordinary R
calls inside duckplyr expressions resolve to Rducks scalar UDFs, without writing
`dd$...` calls. The bridge defaults to row-wise `mode = "scalar"` and can use
`mode = "vectorized"` for vectorized helpers; its marshalling comes from the
current Rducks execution plan.

```{r duckplyr}
demo_con <- DBI::dbConnect(
  duckdb::duckdb(config = list(allow_unsigned_extensions = "true")),
  dbdir = ":memory:"
)
Rducks::rducks_enable(demo_con, threads = "single")
DBI::dbWriteTable(demo_con, "scores", data.frame(
  id = 1:3,
  x = c(2, 21, 34),
  label = c("low", "high", "high")
))

scores <- duckplyr::read_sql_duckdb(
  "SELECT * FROM scores",
  con = demo_con,
  prudence = "stingy"
)
local_score <- function(x, label) {
  as.double(x + if (identical(label, "high")) 100 else 0)
}

with(
  demo_con,
  scores |>
    dplyr::mutate(score = local_score(x, label)) |>
    dplyr::select(id, score) |>
    dplyr::collect(),
  rducks_returns = list(local_score = DOUBLE)
)

Rducks::rducks_release(demo_con)
DBI::dbDisconnect(demo_con, shutdown = TRUE)
```

## Build notes

The source and vendored native dependencies used by `configure` live under
`tools/ext/`. During source-package installation, `configure` writes the generated
artifact to `inst/rducks_extension/build/rducks.duckdb_extension` in the build
tree; after installation the runtime path is
`rducks_extension/build/rducks.duckdb_extension` inside the installed package.
`cleanup` removes only the generated artifact, not the source tree needed by
`R CMD build`. This package-bundled extension layout follows precedents such as
[Rduckhts](https://github.com/RGenomicsETL/duckhts/tree/main/r/Rduckhts). DuckDB
C API headers are refreshed explicitly when the supported DuckDB version changes.

```sh
Rscript tools/fetch_duckdb_headers.R --ref v1.5.2
```

The extension uses DuckDB's
[C extension API](https://github.com/duckdb/duckdb/blob/main/src/include/duckdb/main/capi/header_generation/README.md)
and unstable C extension ABI for connection/runtime access, scalar-function
bind/init/state hooks, dynamic bind-time argument inspection, and selection-vector
helpers. This table is generated from `tools/used_duckdb_unstable_api.R` when
`README.Rmd` is rendered:

```{r unstable-api-table, echo=FALSE, results='asis'}
source("tools/used_duckdb_unstable_api.R")
cat(rducks_used_duckdb_unstable_api_markdown("."))
```

See `docs/BUILD.md` for the build and ABI details.

## References

- [DuckDB extension overview](https://duckdb.org/docs/stable/extensions/overview)
- [DuckDB C extension API notes](https://github.com/duckdb/duckdb/blob/main/src/include/duckdb/main/capi/header_generation/README.md)
- [NNG C library](https://nng.nanomsg.org/)
- [nanonext and mirai](https://mirai.r-lib.org/)
- [mori shared objects for R](https://shikokuchuo.net/mori/)
- [DuckDB Quack extension](https://github.com/duckdb/duckdb-quack)
- [Rduckhts bundled-extension precedent](https://github.com/RGenomicsETL/duckhts/tree/main/r/Rduckhts)

```{r cleanup, include=FALSE}
rducks_release(con)
dbDisconnect(con, shutdown = TRUE)
```
