---
title: "duckplyr Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{duckplyr Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
required <- c("DBI", "dplyr", "duckdb", "duckplyr", "Rducks")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  message("Skipping executable duckplyr vignette because packages are missing: ", paste(missing, collapse = ", "))
  knitr::knit_exit()
}
old_duckplyr_env <- Sys.getenv(
  c("DUCKPLYR_FALLBACK_COLLECT", "DUCKPLYR_FALLBACK_INFO", "DUCKPLYR_FALLBACK_AUTOUPLOAD"),
  unset = NA_character_
)
restore_duckplyr_env <- function() {
  for (name in names(old_duckplyr_env)) {
    if (is.na(old_duckplyr_env[[name]])) {
      Sys.unsetenv(name)
    } else {
      do.call(Sys.setenv, stats::setNames(list(old_duckplyr_env[[name]]), name))
    }
  }
}
Sys.setenv(
  DUCKPLYR_FALLBACK_COLLECT = "0",
  DUCKPLYR_FALLBACK_INFO = "0",
  DUCKPLYR_FALLBACK_AUTOUPLOAD = "0"
)
```

Rducks can make selected ordinary R calls inside a duckplyr pipeline available as
DuckDB scalar UDF calls. The goal is not to emulate dplyr fallback in R; the goal
is to keep the pipeline in DuckDB while DuckDB calls registered Rducks functions
for the operations you explicitly opt into.

## Setup

Use a DuckDB connection with unsigned extension loading enabled and enable
Rducks on that connection. `threads = "single"` is the recommended registration
setting for R-backed functions.

```{r setup}
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(duckdb)
  library(duckplyr)
  library(Rducks)
})

con <- DBI::dbConnect(
  duckdb::duckdb(config = list(allow_unsigned_extensions = "true")),
  dbdir = ":memory:"
)
rducks_enable(con, threads = "single")

input <- data.frame(
  id = 1:6,
  x = as.numeric(c(2, 5, 8, 13, 21, 34)),
  label = c("low", "low", "mid", "mid", "high", "high")
)
DBI::dbWriteTable(con, "duckplyr_scores", input)

scores <- duckplyr::read_sql_duckdb(
  "SELECT * FROM duckplyr_scores",
  con = con,
  prudence = "stingy"
)
```

## Why a bridge is needed

A plain R helper in a duckplyr expression is not automatically a DuckDB SQL
function. With stingy fallback and fallback collection disabled, duckplyr should
fail instead of silently pulling data back to R.

```{r fallback-blocked}
local_score <- function(x, label) {
  bonus <- if (identical(label, "high")) 100 else if (identical(label, "mid")) 10 else 0
  as.double(x + bonus)
}

blocked <- tryCatch({
  scores |>
    mutate(score = local_score(x, label)) |>
    collect()
  FALSE
}, error = function(e) {
  message("fallback blocked: ", conditionMessage(e))
  TRUE
})
stopifnot(isTRUE(blocked))
```

## Register selected R helpers for duckplyr

`rducks_with_duckplyr()` captures the duckplyr expression, registers the named R
helpers as dynamic-argument Rducks scalar UDFs, rewrites matching calls to
DuckDB-function calls, and evaluates the rewritten expression. DuckDB still needs
an explicit return type for every registered helper.

```{r scalar-bridge}
out <- rducks_with_duckplyr(
  con,
  scores |>
    mutate(score = local_score(x, label)) |>
    filter(score >= 100) |>
    select(id, label, score) |>
    arrange(id) |>
    collect(),
  returns = list(local_score = DOUBLE)
)
out
```

The `with.duckdb_connection()` method is equivalent when `rducks_returns` is
supplied:

```{r with-method}
out_with <- with(
  con,
  scores |>
    mutate(score = local_score(x, label)) |>
    filter(score >= 100) |>
    select(id, label, score) |>
    arrange(id) |>
    collect(),
  rducks_returns = list(local_score = DOUBLE)
)
identical(out_with, out)
```

## Why scalar mode is the default

A duckplyr call such as `local_score(x, label)` is translated as a SQL scalar
function call in a relational expression. That SQL surface is row-oriented:
DuckDB sees one logical value for each argument and needs one logical result.
Rducks therefore defaults the bridge to `mode = "scalar"`, which lets ordinary R
helpers be written as row functions.

Rducks scalar-UDF *evaluation mode* is still an implementation choice behind
that SQL scalar function. If a helper is vectorized over whole chunks and returns
a vector of the same length, the duckplyr bridge can register it with
`mode = "vectorized"`:

```{r vectorized-mode}
local_score_vec <- function(x, label) {
  as.double(x + ifelse(label == "high", 100, ifelse(label == "mid", 10, 0)))
}

out_vec <- rducks_with_duckplyr(
  con,
  scores |>
    mutate(score = local_score_vec(x, label)) |>
    filter(score >= 100) |>
    select(id, label, score) |>
    arrange(id) |>
    collect(),
  returns = list(local_score_vec = DOUBLE),
  mode = "vectorized"
)
identical(out_vec, out)
```

The `with()` method exposes the same choice as `rducks_mode`.

## Execution plans: in-process and worker-process

Do not confuse `mode = "scalar"` / `"vectorized"` with the Rducks execution plan.
The mode controls whether the R closure is called per row or per DuckDB chunk.
The execution plan controls the transport (`transport = "inproc"` in the current
R process, or `transport = "ipc"` in worker R processes). The duckplyr bridge
uses the current connection plan at the time it registers helpers.

For example, this executed chunk pins the in-process plan before registering and
evaluating a duckplyr helper:

```{r inproc-plan}
rducks_set_execution_plan(
  con,
  rducks_execution_plan("inproc"),
  threads = 1L,
  external_threads = 1L
)

local_plus_c <- function(x) as.double(x + 1)

rducks_with_duckplyr(
  con,
  scores |>
    mutate(y = local_plus_c(x)) |>
    select(id, y) |>
    arrange(id) |>
    collect(),
  returns = list(local_plus_c = DOUBLE),
  mode = "vectorized"
)
```

Worker-process execution is the same axis: select a `transport = "ipc"` plan
before registering the helper. The high-level duckplyr bridge registers helpers
and evaluates the expression in one call, so it is best for simple runs or for
plans whose registration and execution thread settings are the same. If you need
the full pattern of registering under single-thread DuckDB settings and then
widening `threads` / `external_threads` for a parallel `ipc` query, register the
UDF explicitly with `rducks_register_scalar_udf()` and call it from duckplyr via
duckplyr's `dd$function_name(...)` SQL escape hatch, or wrap that two-phase
pattern in your own helper.

## Cleanup

```{r cleanup}
rducks_release(con)
DBI::dbDisconnect(con, shutdown = TRUE)
restore_duckplyr_env()
```
