---
title: "Execution Plans and IPC Workers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Execution Plans and IPC Workers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Execution plans apply to DuckDB scalar UDFs. They choose where a DuckDB chunk is
evaluated and how it is carried to the R function.

## Supported transports

- `transport = "inproc"` (the default): DuckDB vectors are materialized to SEXPs
  directly in extension C and the R function runs in the current R process.
  DuckDB callbacks may arrive off the recorded R thread; the in-process queued
  backend drains all R work on the recorded R thread, so R API work is never
  concurrent. Maps to the internal `direct_main_queue` engine.
- `transport = "ipc"`: the extension encodes each input chunk to Quack wire
  bytes (a DuckDB `BinarySerializer` `DataChunk` subset), ships it to a
  persistent worker R process over NNG, and decodes the wire-encoded result back
  into DuckDB. Maps to the internal `ipc_nng_pool` engine.

Unsupported combinations fail. Rducks does not silently fall back from one
transport to another.

```{r setup-connection}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")
```

## Select the plan before registration

The default execution plan stored on a connection is used for future scalar-UDF
registrations. Register UDFs under the plan you want to test or deploy.

```{r inproc-plan}
rducks_set_execution_plan(
  con,
  rducks_execution_plan("inproc")
)

rducks_register_scalar_udf(
  con,
  name = "r_plus_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER
)
```

For concurrent in-process execution, set the same plan again with wider DuckDB
thread settings before running queries, so the native runtime backend matches
the UDF metadata being exercised. R work still drains on the recorded R thread.

```{r inproc-concurrent}
rducks_set_execution_plan(
  con,
  rducks_execution_plan("inproc"),
  threads = 4L,
  external_threads = 4L
)
dbGetQuery(con, "SELECT sum(r_plus_one((i % 1000)::INTEGER)) AS total FROM range(20000) t(i)")
```

## Worker-process (`ipc`) plan

`transport = "ipc"` starts or connects to persistent R workers that receive
Quack wire-encoded chunks over NNG. Registration still happens under
single-thread DuckDB settings; widen `threads` / `external_threads` afterwards
for query execution. This vignette uses loopback TCP for the local NNG transport
because it is the most portable choice for executed documentation builds; local
IPC transports such as `"ipc"`, `"unix"`, or Linux `"abstract"` remain available
when supported by the host. Windows documentation builds also use a longer
startup/register timeout because worker process startup can be slower there.

```{r ipc-plan}
ipc_workers <- 1L
ipc_transport <- "tcp"
ipc_timeout <- if (identical(Sys.info()[["sysname"]], "Windows")) 120 else 30

ipc_available <- TRUE
ipc_start_error <- NULL
tryCatch({
  rducks_set_execution_plan(
    con,
    rducks_execution_plan(
      "ipc",
      ipc_workers = ipc_workers,
      ipc_transport = ipc_transport,
      ipc_timeout = ipc_timeout
    ),
    threads = 1L,
    external_threads = 1L
  )

  rducks_register_scalar_udf(
    con,
    name = "r_slow_square",
    fun = function(x) {
      Sys.sleep(0.1)
      x * x
    },
    args = DOUBLE,
    returns = DOUBLE,
    mode = "vectorized",
    side_effects = TRUE
  )

  rducks_set_execution_plan(
    con,
    rducks_execution_plan(
      "ipc",
      ipc_workers = ipc_workers,
      ipc_transport = ipc_transport,
      ipc_timeout = ipc_timeout
    ),
    threads = ipc_workers + 1L,
    external_threads = ipc_workers
  )
}, error = function(e) {
  ipc_available <<- FALSE
  ipc_start_error <<- conditionMessage(e)
  message("IPC worker demo unavailable on this host: ", ipc_start_error)
})
```

Managed startup occurs during registration. Rducks starts local mirai workers,
launches the NNG worker loop, pings each endpoint, then broadcasts the closure,
type metadata, NULL/error policy, packages, and selected globals. If
`ipc_endpoints` is supplied, those endpoints are caller-owned worker processes;
Rducks connects to them but does not stop them. Set `ipc_globals_share = "mori"`
to pass large selected globals to the workers through
[mori](https://shikokuchuo.net/mori/) shared memory instead of serializing them.

## Inspect workers

`rducks_ipc_workers()` lists IPC providers known to the current R process. With
`ping = TRUE`, it also checks whether each endpoint responds.

```{r ipc-workers}
if (isTRUE(ipc_available)) {
  rducks_ipc_workers(con)
  rducks_ipc_workers(con, ping = TRUE, timeout = min(ipc_timeout, 30))
} else {
  data.frame(status = "unavailable", reason = ipc_start_error)
}
```

The result is an R-side provider view: runtime token, provider key, backend,
transport, endpoint, compute name, worker index, task state, and optional ping
status. It is not a DuckDB catalog listing.

## What `rducks_release()` does

`rducks_release(con)` detaches the connection-local Rducks state. It also gives
native code a safe main-thread point to release preserved R objects that had to
be queued by off-main destructors.

If this connection is the last Rducks attachment to the DuckDB runtime,
`rducks_release()` additionally:

1. asks the native extension to close local NNG client pools for that runtime
2. keeps caller-supplied external endpoints alive
3. sends stop requests to Rducks-managed local worker endpoints
4. waits briefly for mirai tasks to resolve and collects resolved tasks
5. tears down the local mirai compute with `mirai::daemons(0, .compute = ...)`
6. unlinks local IPC socket paths
7. removes the provider entry from the process-local store

It does **not** unregister DuckDB catalog functions and does not release closures
still owned by live native catalog metadata. Re-register the same SQL
name/signature to replace a scalar UDF implementation.

Weak-reference finalizers provide best-effort cleanup if a connection object is
garbage-collected, but deterministic code should call `rducks_release(con)`
before `DBI::dbDisconnect(con)`.

```{r cleanup, include=FALSE}
rducks_release(con)
DBI::dbDisconnect(con, shutdown = TRUE)
```
