---
title: "Getting Started with Rducks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Rducks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

`Rducks` is an R package plus a DuckDB extension for registering R functions as
DuckDB SQL functions. The extension owns the SQL catalog objects; R supplies the
closures and explicit type semantics.

The usual workflow is:

1. create a DuckDB connection
2. load and enable the Rducks extension
3. register R-backed SQL functions
4. query them through DuckDB
5. call `rducks_release()` before disconnecting when deterministic cleanup
   matters

## Enable Rducks

```{r enable}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")
```

`threads = "single"` is the safest registration setting. You can select a
scalar-UDF execution plan (`transport = "inproc"` or `transport = "ipc"`) before
registration when you want worker-process evaluation instead of the in-process
default.

## Register a scalar UDF

When `args` is omitted, Rducks registers a dynamic DuckDB varargs UDF. DuckDB
binds concrete argument types at each SQL call site, and Rducks materializes
those values as if the effective signature had been declared explicitly.

```{r scalar-dynamic}
score_udf <- rducks_register_scalar_udf(
  con,
  name = "r_score",
  fun = function(row) {
    bonus <- if (identical(row$label, "high")) 100 else 0
    list(
      score = as.double(row$x + bonus),
      parts = as.double(c(row$x, bonus))
    )
  },
  returns = STRUCT(score = DOUBLE, parts = DOUBLE[]),
  side_effects = TRUE
)

DBI::dbGetQuery(con, "
  WITH input AS (
    SELECT struct_pack(x := x::DOUBLE, label := label) AS payload
    FROM (VALUES (2, 'low'), (21, 'high')) AS t(x, label)
  ), scored AS (
    SELECT r_score(payload) AS result FROM input
  )
  SELECT result.score AS score, result.parts AS parts
  FROM scored
")
```

Use `args = NULL` only for a true zero-argument UDF. Use an explicit `args = ...`
when you want to pin the SQL signature and fail binding of other input types.

## Register an aggregate

Aggregates are separate from scalar-UDF execution plans. The current aggregate
API stores R object state and calls R `update()`, `combine()`, and `finalize()`
callbacks on the recorded R thread.

```{r aggregate}
rducks_register_aggregate(
  con,
  name = "r_sum_i32",
  update = function(state, x) {
    if (is.null(state)) state <- 0L
    as.integer(state + x)
  },
  finalize = function(state) if (is.null(state)) NA_integer_ else state,
  args = INTEGER,
  returns = INTEGER
)

DBI::dbGetQuery(
  con,
  paste(
    "SELECT r_sum_i32(i) AS total",
    "FROM (VALUES (1::INTEGER), (2::INTEGER), (NULL::INTEGER)) t(i)"
  )
)
```

## Register a table function

`rducks_register_table()` infers the SQL argument count from the R function
formals and registers those inputs as DuckDB `ANY`. The R function can return a
finite data frame/list or a `rducks_table_stream()` for scan-time batches.

```{r table-function}
rducks_register_table(
  con,
  name = "r_rows",
  fun = function(n) data.frame(i = seq_len(as.integer(n))),
  chunk_size = 2L
)

DBI::dbGetQuery(con, "SELECT * FROM r_rows(3) ORDER BY i")
```

A streaming table function returns a prototype plus a `next_batch()` callback:

```{r table-stream}
rducks_register_table(
  con,
  name = "r_stream_rows",
  fun = function(n) {
    next_i <- 1L
    limit <- as.integer(n)
    rducks_table_stream(
      prototype = data.frame(i = integer()),
      next_batch = function(batch_size) {
        if (next_i > limit) return(NULL)
        hi <- min(limit, next_i + as.integer(batch_size) - 1L)
        out <- data.frame(i = seq.int(next_i, hi))
        next_i <<- hi + 1L
        out
      }
    )
  },
  chunk_size = 2L
)

DBI::dbGetQuery(con, "SELECT sum(i) AS total FROM r_stream_rows(5)")
```

## Stream query results back to R

`rducks_query_stream()` is for R callers that want native DuckDB result chunks
instead of eager `DBI::dbGetQuery()` materialization. Each `next_batch()`
materializes one DuckDB chunk into a data frame directly from DuckDB vectors.

```{r query-stream}
stream <- rducks_query_stream(
  con,
  "SELECT i, i * i AS sq FROM range(10) tbl(i)"
)

repeat {
  batch <- stream$next_batch()
  if (is.null(batch)) break
  print(batch)
}

stream$close()
```

## Release connection-local state

```{r release}
rducks_release(con)
DBI::dbDisconnect(con, shutdown = TRUE)
```

`rducks_release()` detaches the connection-local Rducks view and, when this is
the last Rducks attachment to the DuckDB runtime, stops Rducks-managed local IPC
workers. It does not unregister DuckDB catalog functions.
