---
title: "Type and Value Semantics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type and Value Semantics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

Rducks keeps SQL type semantics explicit. DuckDB owns binding and execution;
Rducks maps DuckDB values into R values, calls the R function, and writes values
back to DuckDB with the declared result type.

## Function kind, scalar-UDF mode, and execution plan

Three concepts are intentionally separate:

- **DuckDB function kind**: scalar UDF, aggregate function, or table function.
- **Scalar-UDF evaluation mode**: `mode = "scalar"` calls R once per row;
  `mode = "vectorized"` calls R once per DuckDB chunk.
- **Scalar-UDF execution plan**: in-process (`transport = "inproc"`) or
  worker-process (`transport = "ipc"`) evaluation.

Changing a connection's default execution plan affects future scalar-UDF
registrations and the matching native runtime backend; it does not rewrite an
existing scalar UDF to a different transport.

```{r setup-connection}
library(DBI)
library(duckdb)
library(Rducks)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
rducks_enable(con, threads = "single")
```

## Declared descriptors

Rducks descriptors describe DuckDB logical types, including primitive, exact, and
composite values.

```{r descriptors}
primitive <- list(INTEGER, DOUBLE, BOOLEAN, VARCHAR)
exact <- list(UUID, HUGEINT, DECIMAL(18, 4), INTERVAL, BIT)
semi_structured <- list(GEOMETRY, VARIANT)
composite <- list(
  INTEGER[],
  ARRAY(DOUBLE, 3),
  STRUCT(id = INTEGER, label = VARCHAR),
  MAP(VARCHAR, DOUBLE),
  UNION(i = INTEGER, s = VARCHAR)
)
```

Declared scalar-UDF arguments pin the SQL signature:

```{r explicit-signature}
rducks_register_scalar_udf(
  con,
  name = "r_add_one",
  fun = function(x) x + 1L,
  args = INTEGER,
  returns = INTEGER
)
```

Omitting `args` registers a dynamic DuckDB varargs function. At bind time,
DuckDB supplies the concrete logical types for the SQL call, and Rducks uses
those bound types for the same input materialization it would use for explicit
`args`.

```{r dynamic-signature}
rducks_register_scalar_udf(
  con,
  name = "r_payload_label",
  fun = function(payload) paste(payload$label, payload$x, sep = ":"),
  returns = VARCHAR
)

DBI::dbGetQuery(con, "
  SELECT r_payload_label(struct_pack(x := 3::INTEGER, label := 'a')) AS label
")
```

Use `args = NULL` for a true zero-argument UDF.

## NULL handling

`null_handling = "default"` follows DuckDB's default scalar-UDF contract: if a
top-level input is SQL `NULL`, DuckDB produces SQL `NULL` without calling R.

`null_handling = "special"` passes top-level SQL `NULL` inputs through to R as
type-specific missing values so the R function can decide what to return.

```{r null-special}
rducks_register_scalar_udf(
  con,
  name = "r_null_special",
  fun = function(x) if (is.na(x)) 5L else x,
  args = INTEGER,
  returns = INTEGER,
  null_handling = "special"
)

DBI::dbGetQuery(con, "SELECT r_null_special(NULL::INTEGER) AS x")
```

Nested NULLs are part of the nested value. Scalar children usually become typed
`NA` values, while nested composite NULLs become `NULL`.

## Error handling and side effects

`exception_handling = "rethrow"` makes R errors fail the SQL query. Other error
handling modes are explicit choices and should be tested with the declared
return type.

Mark functions with `side_effects = TRUE` when they depend on counters,
randomness, time, I/O, mutation, sleeps, external state, or diagnostics. Without
that flag, DuckDB may treat a scalar UDF as pure enough for ordinary SQL
optimization.

## Runtime reference tables

The package exports compact reference tables so tests and documentation can stay
aligned with the implemented semantics.

```{r reference-tables}
rducks_mode_semantics()[, c("mode", "call_granularity", "input_shape")]

rducks_value_semantics()[
  rducks_value_semantics()$duckdb_type %in% c("INTEGER", "VARCHAR", "GEOMETRY", "VARIANT", "STRUCT"),
  c("duckdb_type", "r_value_class", "special_null_argument")
]

rducks_argument_type_mapping(list(
  INTEGER,
  UUID,
  DECIMAL(10, 2),
  STRUCT(a = INTEGER[])
))
```

```{r cleanup, include=FALSE}
rducks_release(con)
DBI::dbDisconnect(con, shutdown = TRUE)
```
